---
output: html_document
editor_options: 
  chunk_output_type: inline
---

Red Wine Quality Exploratory Data Anaylsis by Cynthia Pottin
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)


```

```{r echo=FALSE, Load_the_Data}
rw <- read.csv('wineQualityReds.csv')

##check to see if all the headers are present and look at the variables
names(rw)


```

Introduction: The red wines data set provided by Udacity is described as  
including 1,599 red wines with 11+ variables. While in CSV format I added wine_id
to help in the analysis if we needed an unique identifier for the wines. In this
analysis we will be looking at the chemical properties and how they influence
the quality of the red wine. In the supporting documentation it describes the
quality was decided by wine experts based on a sample of these wines. 
(https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt)
This analysis will focus on a few of the variables that I predicate will have an
impact on wine quality. As stated in the document about the collect of the data
"volatile acidity: the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste, " therefore I would expect to 
see the quality of the wine to go down as the volatile acidity rises. Also pH, I 
would expect as those wines with a more acidtic pH to be lower on the quality
scale.


First, I would like to look at a summary of the dataset. 

```{r echo=FALSE, Summary}
##looking at a summary of the variables
summary(rw)
```

```{r echo=FALSE, facotr}

##changing quality to a factor
rw$quality.f <- factor(rw$quality, levels = c('3', '4', '5', '6', '7', '8', '9'))

```

```{r echo=FALSE, structure}

##looking at the structure of the data set
str(rw)
```

Looking at the structure of the data. Quality is an int type and we will need it
to be a factor for use as a categorical data later in our charting.

Quality is the first variable and the one of most interest to look at. It appears
that the quality ranges from 3.000 to 8.000.

##Univariate Plots Section

```{r echo=FALSE, Univariate_Plots_1}

##plotting bar graph of the quality and the count of each wine in those categories
qplot(data = rw, x = quality.f, fill = I('#b30000'))


```

This histogram gives a look at the count of the wines that have been grouped
into each of the quality numeric classifications. It appears that the majority
of the red wines are a 5 in quality, but let's look at a summary of that 
information as well. 

```{r echo=FALSE, Univariate_analysis_tableq}
##this creates a table that will let one see the numeric count of the wines for each rating
table(rw$quality)

```
This table shows that 681 wines are in the 5 classification and 638 are in the 
6 classification of quality. This shows that most wines call in the middle of
the quality ratings. 

```{r echo=FALSE, Univariate_Plots_pH}
##histogram of pH
qplot(data = rw, x = pH, fill = I('#b30000'), binwidth = 0.1)


```

This histogram of pH has a fairly normal distribution shape. 


```{r echo=FALSE, Univariate_Plots_VA}
##histogram of volatile acidity
qplot(data = rw, x = volatile.acidity, fill = I('#b30000'), binwidth = 0.05)

```

This histogram has two peak, one around 0.4 and the other around 0.6.

```{r echo=FALSE, Univariate_Plots_FA}
##histogram of fixed acidity
qplot(data = rw, x = fixed.acidity, fill = I('#b30000'), binwidth = 1)

```

This histogram is skewed to the left.

```{r echo=FALSE, CA}
##histogram of citric acid
qplot(data = rw, x = citric.acid, fill = I('#b30000'), binwidth = 0.05)

```

Citric Acid is a bimodial histogram

```{r echo=FALSE, RS}
##histogram of the residual sugar
qplot(data = rw, x = residual.sugar, fill = I('#b30000'), binwidth = 0.25)

```

This histogram has a high spike around 2 and is positively skewed. Most of the
wines appear to have residual sugar of less than 4. 

```{r echo=FALSE, Chlorides}
##historgram of the chlorides
qplot(data = rw, x = chlorides, fill = I('#b30000'), binwidth = 0.01)

```

Most of the results are clustered between 0 and 0.1 however we have a few that
fall way outside of that as well. This leads me to believe that chlorides will 
have no correlation to quality rating. 

```{r echo=FALSE, FSD}
##histogram of the free sulfur dioxide
qplot(data = rw, x = free.sulfur.dioxide, fill = I('#b30000'), binwidth = 1)

```

This histogram is skewed roght as well with a few outliners outside of 60.

```{r echo=FALSE, TSD}
##histogram total sulfur dioxide
qplot(data = rw, x = total.sulfur.dioxide, fill = I('#b30000'), binwidth = 5)

```

Skewed right again with a few closer to 300. 

```{r echo=FALSE, Density}
##histogram of density
qplot(data = rw, x = density, fill = I('#b30000'), binwidth = 0.001)

```

Density has a normal distribution on this histogram.

```{r echo=FALSE, sulphates}
##histogram of sulphates
qplot(data = rw, x = sulphates, fill = I('#b30000'), binwidth = 0.05)

```


```{r echo=FALSE, alcohol}
##histogram of alcohol
qplot(data = rw, x = alcohol, fill = I('#b30000'), binwidth = 0.1)

```

##Univariate Analysis: 

When looking at the red wine data in the univariate 
analysis the main focus is to look at the variables and their distribution in the
wines in the dataset. The main one of interest is quality, but it is important
to look at the other variables and see if we can start to see which variables 
will most heavily impact the quality rating of the wines. 

I would think that acidity will cause a wine to be rated lower, so I will be
looking at the pH, volatile acidity and the fixed acidity. I would also 
anticipate that wines with lots of residual sugars will be rated lower due 
to their sweetness.

I added wine_id to my dataset in cause we need a unique identifier for the wine.
I have not preformed any functions to tidy or adjust the data. 
Looking over these histograms, 

Also, did some additional reading and research on wine and they add sulfur
dioxide to the wine. In the article, it was discussed that this may have an
impact on the flavor. I will be looking at the sulphates and sulfur dioxide
variables as well. 

##Bivariate Plots Section

In this next section, I would like to look at the relationship between acidity, 
sugar, alcholol and quality.



```{r echo=FALSE, PC_QVA}
##calculation of pearson's correlation coefficient between quality and volatile.acidity
with(rw, cor.test(quality, volatile.acidity, method = 'pearson'))

```

This shows a medium negative correlation between quality and volatile acidity.

```{r echo=FALSE, Bivariate_Plots_VAQ}
##plot of the scatter plot for 2 variables
##geom_smooth aids the eye in seeing the patterns with a line

ggplot(aes(y = volatile.acidity, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
  

```

Scatter plot of quality and volatile acidity, showing a medium negative 
correlation. 


```{r echo=FALSE, PC_FAQ}
with(rw, cor.test(quality, fixed.acidity, method = 'pearson'))
```


```{r echo=FALSE, Bivariate_Plots_FAQ}

##plot of the scatter plot for 2 variables
##geom_smooth aids the eye in seeing the patterns with a line
ggplot(aes(y = fixed.acidity, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)

```


These graphs and the correlation shows a small positive correlation for quality 
and fixed acidity.

```{r echo=FALSE, PC_AQ}
with(rw, cor.test(quality, alcohol, method = 'pearson'))
```


Rounding this to 0.5 shows a large positive correlation to alcohol and quality.

```{r echo=FALSE, Bivariate_Plots_AQ}

##plot of the scatter plot for 2 variables
##geom_smooth aids the eye in seeing the patterns with a line

ggplot(aes(y = alcohol, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

This shows a large positive correlation to alcohol and quality. 
It appears that higher quality wines have a higher alcohol cotent. 


```{r echo=FALSE, PC_QS}
with(rw, cor.test(quality, sulphates, method = 'pearson'))
```

```{r echo=FALSE, Bivariate_Plots_SQ}
ggplot(aes(y = sulphates, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

```{r echo=FALSE, PC_qph}
with(rw, cor.test(quality, pH, method = 'pearson'))
```

```{r echo=FALSE, Bivariate_Plots_QpH}
ggplot(aes(y = pH, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

pH seems to have very little correlation to quality rating.

```{r echo=FALSE, PC_QRS}
with(rw, cor.test(quality, residual.sugar, method = 'pearson'))
```

```{r echo=FALSE, Bivariate_Plots_QRS}
ggplot(aes(y = residual.sugar, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

```{r echo=FALSE, PC_SQ}
with(rw, cor.test(quality, sulphates, method = 'pearson'))
```


```{r echo=FALSE, Bivariate_Plots_SulQual}
ggplot(aes(y = sulphates, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

```{r echo=FALSE, PC_QD}
with(rw, cor.test(quality, density, method = 'pearson'))
```

```{r echo=FALSE, Bivariate_Plots_DQ}
ggplot(aes(y = density, x = quality), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

```{r echo=FALSE, PC_DFA}
with(rw, cor.test(density, fixed.acidity, method = 'pearson'))
```

```{r echo=FALSE, Bivariate_Plots_DFA}
ggplot(aes(y = density, x = fixed.acidity), data = rw) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method=lm)
```

As density increases fixed acidity increases as well. This shows a strong
correlation, but not one that I would have expected. I would expected a stronger
correlation to the residual sugars and density. 

##Bivariate Analysis

It was observed that alcohol content and quality have a positive correlation and
volatile acidity and quality have a negative correlation. These two factors seem
to impact the quality rating of the wines. 


There appears to be a strong correlation between density and fixed acidity.
Looking that the data I would have thought that the residual sugar would have
a stronger correlation to density than fixed acidity.

Strongest relationship in realtion to quality found was the alcohol content.



##Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots_QVApH}
rw$quality.f <- factor(rw$quality, levels = c('3', '4', '5', '6', '7', '8', '9'))

ggplot(aes(x = volatile.acidity, y = pH, colour = quality.f), data = rw) + 
  geom_point(alpha = 0.5, size = 1,  position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(limits = c(0.2, 1),
                     breaks = c(0.2, 0.4, 0.6, 0.8, 1)) + 
  scale_y_continuous(limits = c(2.5, 4.5),
                     breaks = c(2.5, 3, 3.5, 4, 4.5, 5)) +
  ggtitle('pH vs. Volatile Acidity')

```

This shows that most of the higher quality wines have a lower Volatile Acidity
level. Most of the wines have a pH better 3 and 3.5 so that doesn't seem to be
as large of an impoact on the quality rating. 

```{r echo=FALSE, Multivariate_Plots_AVAQ}


ggplot(aes(x = volatile.acidity, y = alcohol, colour = quality.f), data = rw) + 
  geom_point(alpha = 0.5, size = 1,  position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(limits = c(0, 1.6),
                     breaks = c(0, 0.5, 1.0 , 1.5, 2.0)) + 
  scale_y_continuous(limits = c(8, 15),
                     breaks = c(8, 9, 10, 11, 12, 13, 14, 15)) +
  ggtitle('Alcohol vs. Volatile Acidity')

```

This chart shows that lower volatile acidity and higher alcohol content is
present in the higher quality wines. 

```{r echo=FALSE, Multivariate_Plots_DAQF}


ggplot(aes(x = density, y = alcohol, colour = quality.f), data = rw) + 
  geom_point(alpha = 0.5, size = 1,  position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(limits = c(0.99, 1),
                     breaks = c(0.99, 0.9925, 0.995, 0.9975, 1.0)) + 
  scale_y_continuous(limits = c(8, 15),
                     breaks = c(8, 9, 10, 11, 12, 13, 14, 15)) +
  ggtitle('Alcohol vs. Density')

```

```{r echo=FALSE, Multivariate_Plots_VASQF}


ggplot(aes(x = volatile.acidity, y = sulphates, colour = quality.f), data = rw) + 
  geom_point(alpha = 0.5, size = 1,  position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(limits = c(0, 1.6),
                     breaks = c(0, 0.5, 1.0 , 1.5, 2.0)) + 
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  ggtitle('Sulphates vs. Volatile Acidity')

```
The chart shows that wines in the higher quality ratings have a lower volatile
acidity and a higher sulphate content. 


##Multivariate Analysis

These were great for refining the relationships seen in the other charts. 
Wines in the higher quality ratings have a lower volatile
acidity and a higher sulphate content. 
Wines with lower volatile acidity and higher alcohol content have a higher 
quality rating.

------

##Final Plots and Summary


Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = quality.f), data = rw) +
  geom_bar(stat = 'Count', fill = I('#b30000')) +
  xlab('Quality') +
  ggtitle('Distribution of Wine Qualities') + 
  theme(text = element_text(size = 10))


```

This the is overall distribution of the wine qualities for the red wine data set. 
As you can see the majority of wines fall into the rating of 5 and 6.

Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x = volatile.acidity, y = alcohol, colour = quality.f), data = rw) + 
  geom_point(alpha = 0.5, size = 1,  position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(limits = c(0, 1.6),
                     breaks = c(0, 0.5, 1.0 , 1.5, 2.0)) + 
  scale_y_continuous(limits = c(8, 15),
                     breaks = c(8, 9, 10, 11, 12, 13, 14, 15)) +
  xlab('Volatile Acidity (g / dm^3)') +
  ylab('Alcohol (% by volume)') +
  ggtitle('Alcohol vs. Volatile Acidity')

```

This chart compared Volatile Acidity, Alcohol content and Quality rate of the 
red wine data set. It shows the some clustering of the higher quality wines in 
the area of lower volatile acidity and higher alcohol content. 


Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = volatile.acidity, y = sulphates, colour = quality.f), data = rw) + 
  geom_point(alpha = 0.5, size = 1,  position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(limits = c(0, 1.6),
                     breaks = c(0, 0.5, 1.0 , 1.5, 2.0)) + 
  scale_y_continuous(limits = c(0.25, 1),
                     breaks = c(0.25, 0.5, 0.75, 1)) +
  xlab('Volatile Acidity (g / dm^3)') +
  ylab('Sulphates (g / dm3)') +
  ggtitle('Sulphates vs. Volatile Acidity')

```

This chart compared Volatile Acidity, Sulphates content and Quality rate of the 
red wine data set. It shows the some clustering of the higher quality wines in 
the area of lower volatile acidity and higher sulphates present in the wine.

------

##Reflection

In analyzing this data, I have discovered that volatile acidity has a negative
impact on wine rating. This doesn't come as a surprise since through my 
pre-reading for this project they mentioned that higher volatile acidity can make 
the wine taste more like vinegar. 
I think it was a surprise that alcohol content wines had the strongest
positive correlation to the quality rating. I would like the alcohol content 
would have no impact on the flavor and therefore no impact on the rating. 
The sulphates present and the correlation to the wine rating was the most 
interesting to me, since I did not
go into this project thinking sulphates with would impact the rating. 
I think more data about the temperature of the storage, types of grapes, types 
of barrels, and added ingredients would have been interesting to explore as well. 
I did struggle with the subjective nature of the rating for the wine. I feel
that we could use this data coupled with other data points to possible 
eliminate the need for the human element and thus the subject rating.

One thing that was very helpful for me in this course was as I worked through
the lessons, I would work with not only the data set for the lesson, but also
my project data set. This was extremely helpful to start thinking about the 
data as I worked through the course.


References:
http://www.sthda.com/english/wiki/ggplot2-scatter-plots-quick-start-guide-r-software-and-data-visualization

https://statistics.laerd.com/statistical-guides/pearson-correlation-coefficient-statistical-guide.php

https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt

https://www.torres.es/en/blog/how-wine-made/4-factors-determine-wine-quality#

https://swcarpentry.github.io/r-novice-inflammation/12-supp-factors/index.html
